---
title: |
    | Bayesian statistics for ecology 
    | 7. Contrast scientific hypotheses with model selection
author: "Olivier Gimenez"
date: "March 2021"
output:
  beamer_presentation:
    fig_caption: no
    includes:
      in_header: header.tex
    latex_engine: pdflatex
    slide_level: 2
    theme: metropolis
  ioslides_presentation: default
classoption: aspectratio=169
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = FALSE, 
                      echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      fig.height=6, 
                      fig.width = 1.777777*6,
                      tidy = FALSE, 
                      comment = NA, 
                      highlight = TRUE, 
                      prompt = FALSE, 
                      crop = TRUE,
                      comment = "#>",
                      collapse = TRUE)
knitr::opts_knit$set(width = 60)
library(tidyverse)
library(reshape2)
library(R2jags)
theme_set(theme_light(base_size = 16))
make_latex_decorator <- function(output, otherwise) {
  function() {
      if (knitr:::is_latex_output()) output else otherwise
  }
}
insert_pause <- make_latex_decorator(". . .", "\n")
insert_slide_break <- make_latex_decorator("----", "\n")
insert_inc_bullet <- make_latex_decorator("> *", "*")
insert_html_math <- make_latex_decorator("", "$$")
```


# Model selection

## How to select a best model?

`r insert_inc_bullet()` Is there any effect of rain or temperature or both on breeding success?

`r insert_inc_bullet()` The proportion of explained variance $R^2$ is problematic, because the more variables you have, the bigger $R^2$ is.

`r insert_inc_bullet()` Idea: **penalize models with too many parameters**.

## Akaike information criterion (AIC)

$$AIC = - 2 \log(L(\hat{\theta}_1,\ldots,\hat{\theta}_K)) + 2 K$$

with $L$ the likelihood and $K$ the number of parameters $\theta_i$.

## Akaike information criterion (AIC)

$$\text{AIC} = {\color{red}{- 2 \log(L(\hat{\theta}_1,\ldots,\hat{\theta}_K))}} + 2 K$$

\textcolor{red}{A measure of goodness-of-fit of the model to the data}: the more parameters you have, the smaller the deviance is (or the bigger the likelihood is).

## Akaike information criterion (AIC)

$$\text{AIC} = - 2 \log(L(\hat{\theta}_1,\ldots,\hat{\theta}_K)) + {\color{red}{2 K}}$$

\textcolor{red}{A penalty}: twice the number of parameters $K$

## Akaike information criterion (AIC)

`r insert_inc_bullet()` AIC makes the balance between *quality of fit* and *complexity* of a model.

`r insert_inc_bullet()` Best model is the one with lowest AIC value.

`r insert_inc_bullet()` Two models are difficult to distinguish if $\Delta \text{AIC} < 2$.

## Bayesian version

* Watanabe-Akaike Information Criteria or WAIC:

$$\textrm{WAIC} = -2 \sum_{i = 1}^n \log E[p(y_i \mid \theta)] + 
                  2 p_\text{WAIC}$$

* where $E[p(y_i \mid \theta)]$ is the posterior mean of the likelihood of the $i$th observation and

* $p_\text{WAIC}$ is the effective number of parameters computed using the posterior variance of the likelihood. 

* Relatively new and not yet available in Jags in routine.

```{r include = FALSE}
nbchicks <- c(151,105,73,107,113,87,77,108,118,122,112,120,122,89,69,71,
              53,41,53,31,35,14,18)

nbpairs <- c(173,164,103,113,122,112,98,121,132,136,133,137,145,117,90,80,
            67,54,58,39,42,23,23)

temp <- c(15.1,13.3,15.3,13.3,14.6,15.6,13.1,13.1,15.0,11.7,15.3,14.4,14.4,
         12.7,11.7,11.9,15.9,13.4,14.0,13.9,12.9,15.1,13.0)

rain <- c(67,52,88,61,32,36,72,43,92,32,86,28,57,55,66,26,28,96,48,90,86,
           78,87)

datax <- list(N = 23, nbchicks = nbchicks, nbpairs = nbpairs, 
              temp = (temp - mean(temp))/sd(temp), 
              rain = (rain - mean(rain))/sd(rain))

```


## WAIC in Jags

```{r include=FALSE}
model <- 
paste("
model
{
	for( i in 1 : N) 
		{
		nbchicks[i] ~ dbin(p[i],nbpairs[i])
		logit(p[i]) <- a + b.temp * temp[i] + b.rain * rain[i]
		}
			
# priors for regression parameters
a ~ dnorm(0,0.001)
b.temp ~ dnorm(0,0.001)
b.rain ~ dnorm(0,0.001)
			
	}
")
writeLines(model,"code/logistic.txt")
init1 <- list(a = -0.5, b.temp = -0.5, b.rain = -0.5)
init2 <- list(a = 0.5, b.temp = 0.5, b.rain = 0.5)
inits <- list(init1,init2)
parameters <- c("a","b.temp","b.rain")
nb.burnin <- 1000
nb.iterations <- 2000
storks <- jags(data  = datax,
               inits = inits,
               parameters.to.save = parameters,
               model.file = "code/logistic.txt",
               n.chains = 2,
               n.iter = nb.iterations,
               n.burnin = nb.burnin)
```

```{r}
# calculate wAIC with JAGS
# https://sourceforge.net/p/mcmc-jags/discussion/610036/thread/8211df61/#ea5c
samples <- jags.samples(storks$model,c("WAIC","deviance"), type = "mean", 
						n.iter = 2000,
						n.burnin = 1000,
						n.thin = 1)
```

## WAIC in Jags

```{r}
samples$p_waic <- samples$WAIC
samples$waic <- samples$deviance + samples$p_waic
tmp <- sapply(samples, sum)
waic <- round(c(waic = tmp[["waic"]], p_waic = tmp[["p_waic"]]),1)
waic
```

# Your turn

## Model selection with WAIC

* Fit models with rainfall effect, temperature effect and without any covariate.

* Rank them with WAIC.

# Solution

## Model with temperature only

```{r}
# model specification
model <- 
paste("
model
{
	for( i in 1 : N) 
		{
		nbchicks[i] ~ dbin(p[i],nbpairs[i])
		logit(p[i]) <- a + b * cov[i]
		}
			
# priors for regression parameters
a ~ dnorm(0,0.001)
b ~ dnorm(0,0.001)
	}
")
writeLines(model,"code/logtemp.txt")
```

`r insert_slide_break()`

```{r}
# list of lists of initial values (one for each MCMC chain)
init1 <- list(a = -0.5, b = -0.5)
init2 <- list(a = 0.5, b = 0.5)
inits <- list(init1,init2)
# specify parameters that need to be estimated
parameters <- c("a","b")
# specify nb iterations for burn-in and final inference 
nb.burnin <- 1000
nb.iterations <- 2000
# read in data
datax <- list(N = 23, nbchicks = nbchicks, nbpairs = nbpairs, 
              cov = (temp - mean(temp))/sd(temp))
```

`r insert_slide_break()`

```{r message=FALSE, warning=FALSE}
# load R2jags to run Jags through R
storks_temp <- jags(data  = datax,
               inits = inits,
               parameters.to.save = parameters,
               model.file = "code/logtemp.txt",
               n.chains = 2,
               n.iter = nb.iterations,
               n.burnin = nb.burnin)
```

`r insert_slide_break()`

```{r}
# compute WAIC
samples <- jags.samples(storks_temp$model,c("WAIC","deviance"), type = "mean", 
						n.iter = 2000,
						n.burnin = 1000,
						n.thin = 1)
samples$p_waic <- samples$WAIC
samples$waic <- samples$deviance + samples$p_waic
tmp <- sapply(samples, sum)
waic_temp <- round(c(waic = tmp[["waic"]], p_waic = tmp[["p_waic"]]),1)
```

## Model with rainfall only

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# read in data
datax <- list(N = 23, nbchicks = nbchicks, nbpairs = nbpairs, 
              cov = (rain - mean(rain))/sd(rain))
```

`r insert_slide_break()`

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# load R2jags to run Jags through R
storks_temp <- jags(data  = datax,
               inits = inits,
               parameters.to.save = parameters,
               model.file = "code/logtemp.txt",
               n.chains = 2,
               n.iter = nb.iterations,
               n.burnin = nb.burnin)
```

`r insert_slide_break()`

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# compute WAIC
samples <- jags.samples(storks_temp$model,c("WAIC","deviance"), type = "mean", 
						n.iter = 2000,
						n.burnin = 1000,
						n.thin = 1)
samples$p_waic <- samples$WAIC
samples$waic <- samples$deviance + samples$p_waic
tmp <- sapply(samples, sum)
waic_rain <- round(c(waic = tmp[["waic"]], p_waic = tmp[["p_waic"]]),1)
```

## Model with no effect of covariates

```{r}
# model specification
model <- 
paste("
model
{
	for( i in 1 : N) 
		{
		nbchicks[i] ~ dbin(p[i],nbpairs[i])
		logit(p[i]) <- a
		}
			
# priors for regression parameters
a ~ dnorm(0,0.001)
	}
")
writeLines(model,"code/lognull.txt")
```

`r insert_slide_break()`

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# list of lists of initial values (one for each MCMC chain)
init1 <- list(a = -0.5)
init2 <- list(a = 0.5)
inits <- list(init1,init2)
# specify parameters that need to be estimated
parameters <- c("a")
# specify nb iterations for burn-in and final inference 
nb.burnin <- 1000
nb.iterations <- 2000
# read in data
datax <- list(N = 23, nbchicks = nbchicks, nbpairs = nbpairs)
```

`r insert_slide_break()`

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# load R2jags to run Jags through R
storks_temp <- jags(data  = datax,
               inits = inits,
               parameters.to.save = parameters,
               model.file = "code/lognull.txt",
               n.chains = 2,
               n.iter = nb.iterations,
               n.burnin = nb.burnin)
```

`r insert_slide_break()`

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# compute WAIC
samples <- jags.samples(storks_temp$model,c("WAIC","deviance"), type = "mean", 
						n.iter = 2000,
						n.burnin = 1000,
						n.thin = 1)
samples$p_waic <- samples$WAIC
samples$waic <- samples$deviance + samples$p_waic
tmp <- sapply(samples, sum)
waic_null <- round(c(waic = tmp[["waic"]], p_waic = tmp[["p_waic"]]),1)
```

## Compare WAIC

```{r}
data.frame(model = c('both_covariates', 'temp', 'rain', 'none'),
           waic = c(waic[1],waic_temp[1],waic_rain[1],waic_null[1]),
           p_waic = c(waic[2],waic_temp[2],waic_rain[2],waic_null[2])) %>%
  arrange(waic)
```

Model with rainfall only seems to be better supported by the data. In case models have similar WAIC values, model-averaging might be useful. 
