---
title: "Practical 4"
author: "Olivier Gimenez"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE}
#options(htmltools.dir.version = FALSE)
#knitr::opts_chunk$set(comment = "", warning = FALSE, message = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(palmerpenguins)
library(tidyverse)
```

## Prior predictive checks

* Make sure you can reproduce the figures of the corresponding lecture slides

* Consider a linear regression with unreasonable and reasonable priors

* Consider a logistic regression with unreasonable and reasonable priors

---
## Solutions

---
## Linear regression

* Unreasonable prior $\beta \sim N(0, 1000^2)$

.center[
```{r, echo = FALSE, fig.width = 5.5, fig.asp = 0.8, dev = "svg"}
plot(density(rnorm(1000, 0, 1000)),   
     main="", xlab="Height (m)")
```
]

---
## Linear regression

* Reasonable prior $\beta \sim N(2, 0.5^2)$

.center[
```{r, echo = FALSE, fig.width = 5.5, fig.asp = 0.8, dev = "svg"}
plot(density(rnorm(1000, 2, 0.5)),   
      main="", xlab="Height (m)")
```
]

---
## Logistic regression

* Unreasonable prior $\text{logit}(\phi) = \beta \sim N(0, 10^2)$

.center[
```{r, echo = FALSE, fig.width = 5.5, fig.asp = 0.8, dev = "svg"}
plot(density(plogis(rnorm(1000,0,10)), 
from = 0, to = 1), main='', xlab='survival')
```
]

---
## Logistic regression

* Reasonable prior $\text{logit}(\phi) = \beta \sim N(0, 1.5^2)$

.center[
```{r, echo = FALSE, fig.width = 5.5, fig.asp = 0.8, dev = "svg"}
plot(density(plogis(rnorm(1000,0,1.5)), 
from = 0, to = 1), main='', xlab='survival')
```
]