---
title: "Practical 2"
author: "Olivier Gimenez"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE}
#options(htmltools.dir.version = FALSE)
#knitr::opts_chunk$set(comment = "", warning = FALSE, message = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(palmerpenguins)
library(tidyverse)
```

## MLE of the parameters of a Normal distribution

* Assume we have collected data on the height of 100 people:
```{r}
# set seed for random numbers
set.seed(2020)
# simulate data from Normal distribution
n <- 100
height <- rnorm(n, mean=170, sd=10)
```

---

.center[
```{r echo=FALSE, fig.width = 5.5, fig.asp = 1, dev = "svg"}
data.frame(height = height) %>%
  ggplot(aes(x = height)) + 
  geom_histogram(color = "blue",
                 fill = "dodgerblue") + 
  labs(x = "Height", 
       y = "Density") +
  theme_minimal(base_size = 20)
```
]

---

* We consider a Normal distribution for the model. 

* Compute the MLE of the parameters of the Normal distribution.

* Hint: Use functions `optim()` and `dnorm()`

---

# Solution

Function for likelihood of a normal distribution with mean $\mu$ and standard deviation $\sigma$:
```{r}
negloglik <- function(theta, data) {
  mu <- theta[1]
  sigma <- theta[2]
  x <- data
  -sum(dnorm(x, mean = mu, sd = sigma, log = TRUE))
}
negloglik(theta = c(150,1), height)
```

---

* Minimiiiiiize
```{r}
fit <- optim(par = c(1,1), fn = negloglik, data = height)
fit
```


---

.center[
```{r, echo = FALSE, fig.width = 5.5, fig.asp = 1, dev = "svg"}
binwidth <- 1 
df <- data.frame(x = height) %>%
  ggplot(aes(x = height, 
             mean = fit$par[1], 
             sd = fit$par[2], 
             binwidth = binwidth, 
             n = n)) +
  geom_histogram(binwidth = binwidth, 
                 colour = "white", 
                 fill = "cornflowerblue", 
                 size = 0.1) +
  stat_function(fun = function(x) dnorm(x = x, 
                                        mean = fit$par[1], 
                                        sd = fit$par[2]) * n * binwidth,
                color = "darkred", 
                size = 1) +
  theme_minimal(base_size = 20)
df
```
]

