---
title: "Practical 6"
author: "Olivier Gimenez"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE}
#options(htmltools.dir.version = FALSE)
#knitr::opts_chunk$set(comment = "", warning = FALSE, message = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse)
```

## Working with the MCMC outputs

* From the values you generated in the posterior distribution of parameter say $\theta$, you may get the posterior distribution of any function of $\theta$

* For example, back to the stork Bayesian analysis, get the posterior distribution of $b^2_{rain} + \cos(b_{temp})$

---
## Solutions

```{r include = FALSE, echo = FALSE}
nbchicks <- c(151,105,73,107,113,87,77,108,118,122,112,120,122,89,69,71,
              53,41,53,31,35,14,18)

nbpairs <- c(173,164,103,113,122,112,98,121,132,136,133,137,145,117,90,80,
            67,54,58,39,42,23,23)

temp <- c(15.1,13.3,15.3,13.3,14.6,15.6,13.1,13.1,15.0,11.7,15.3,14.4,14.4,
         12.7,11.7,11.9,15.9,13.4,14.0,13.9,12.9,15.1,13.0)

rain <- c(67,52,88,61,32,36,72,43,92,32,86,28,57,55,66,26,28,96,48,90,86,
           78,87)

datax <- list(N = 23, nbchicks = nbchicks, nbpairs = nbpairs, 
              temp = (temp - mean(temp))/sd(temp), 
              rain = (rain - mean(rain))/sd(rain))

# list of lists of initial values (one for each MCMC chain)
init1 <- list(a = -0.5, b.temp = -0.5, b.rain = -0.5)
init2 <- list(a = 0.5, b.temp = 0.5, b.rain = 0.5)
inits <- list(init1,init2)

# specify parameters that need to be estimated
parameters <- c("a","b.temp","b.rain")

# specify nb iterations for burn-in and final inference 
nb.burnin <- 1000
nb.iterations <- 2000

# load R2jags
library(R2jags)
# run Jags
storks <- jags(data  = datax,
               inits = inits,
               parameters.to.save = parameters,
               model.file = here::here("slides", "code","logistic.txt"),
               n.chains = 2,
               n.iter = nb.iterations,
               n.burnin = nb.burnin)

res <- as.mcmc(storks) # convert outputs in a list
res <- rbind(res[[1]],res[[2]]) # put two MCMC lists on top of each other

```

---

* Evaluate the function for each MCMC iteration
```{r}
stupid <- res[,'b.rain']^2 + cos(res[,'b.temp'])
```

.tiny-font[
```{r}
stupid
```
]

---

* Plot the distribution 1/2
.center[
```{r, echo = FALSE, fig.width = 5.5, fig.asp = 0.8, dev = "svg"}
plot(density(stupid), xlab = '', main = '', lwd = 3)
```
]

---

* Plot the distribution 2/2
.center[
```{r, echo = FALSE, fig.width = 5.5, fig.asp = 0.8, dev = "svg"}
hist(stupid)
```
]

---

* Get the probability that this quantity is strictly positive
```{r}
mean(stupid > 0)
```



