---
title: "Practical 7"
author: "Olivier Gimenez"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE}
#options(htmltools.dir.version = FALSE)
#knitr::opts_chunk$set(comment = "", warning = FALSE, message = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(R2jags)
library(tidyverse)
```

## Model selection with wAIC

* Fit models with rainfall effect, temperature effect and without any covariate to the stork data.

* Rank them with wAIC.

---
# Solution

---
## Data

```{r }
nbchicks <- c(151,105,73,107,113,87,77,108,118,122,112,120,122,
              89,69,71,53,41,53,31,35,14,18)

nbpairs <- c(173,164,103,113,122,112,98,121,132,136,133,137,
             145,117,90,80,67,54,58,39,42,23,23)

temp <- c(15.1,13.3,15.3,13.3,14.6,15.6,13.1,13.1,15.0,11.7,
          15.3,14.4,14.4,12.7,11.7,11.9,15.9,13.4,14.0,13.9,
          12.9,15.1,13.0)

rain <- c(67,52,88,61,32,36,72,43,92,32,86,28,57,55,66,26,
          28,96,48,90,86,78,87)
```

---
## Model with both covariates

.tiny-font[
```{r}
model <- 
paste("
model
{
	for( i in 1 : N) 
		{
		nbchicks[i] ~ dbin(p[i],nbpairs[i])
		logit(p[i]) <- a + b.temp * temp[i] + b.rain * rain[i]
		}
			
# priors for regression parameters
a ~ dnorm(0,0.001)
b.temp ~ dnorm(0,0.001)
b.rain ~ dnorm(0,0.001)
			
	}
")
writeLines(model,here::here("slides", "code", "logistic.txt"))
```
]

---
## MCMC details and data

```{r}
init1 <- list(a = -0.5, b.temp = -0.5, b.rain = -0.5)
init2 <- list(a = 0.5, b.temp = 0.5, b.rain = 0.5)
inits <- list(init1,init2)
parameters <- c("a","b.temp","b.rain")
nb.burnin <- 1000
nb.iterations <- 2000

datax <- list(N = 23, 
              nbchicks = nbchicks, 
              nbpairs = nbpairs, 
              temp = (temp - mean(temp)) / sd(temp), 
              rain = (rain - mean(rain)) / sd(rain))
```

---
## Run Jags

.tiny-font[
```{r}
storks <- jags(data  = datax,
               inits = inits,
               parameters.to.save = parameters,
               model.file = here::here("slides", "code", "logistic.txt"),
               n.chains = 2,
               n.iter = nb.iterations,
               n.burnin = nb.burnin)
```
]

---
## Calculate wAIC with Jags

* Check out post at <https://sourceforge.net/p/mcmc-jags/discussion/610036/thread/8211df61/#ea5c>

.tiny-font[
```{r}
samples <- jags.samples(storks$model,c("WAIC","deviance"), type = "mean", 
						n.iter = 2000,
						n.burnin = 1000,
						n.thin = 1)
samples$p_waic <- samples$WAIC
samples$waic <- samples$deviance + samples$p_waic
tmp <- sapply(samples, sum)
waic <- round(c(waic = tmp[["waic"]], 
                p_waic = tmp[["p_waic"]]),1)
waic
```
]

---
## Model with temperature only

.tiny-font[
```{r}
# model specification
model <- 
paste("
model
{
	for( i in 1 : N) 
		{
		nbchicks[i] ~ dbin(p[i],nbpairs[i])
		logit(p[i]) <- a + b * cov[i]
		}
			
# priors for regression parameters
a ~ dnorm(0,0.001)
b ~ dnorm(0,0.001)
	}
")
writeLines(model,here::here("slides", "code", "logtemp.txt"))
```
]

---
## MCMC details and data

.small-font[
```{r}
# list of lists of initial values (one for each MCMC chain)
init1 <- list(a = -0.5, b = -0.5)
init2 <- list(a = 0.5, b = 0.5)
inits <- list(init1,init2)
# specify parameters that need to be estimated
parameters <- c("a","b")
# specify nb iterations for burn-in and final inference 
nb.burnin <- 1000
nb.iterations <- 2000

# read in data
datax <- list(N = 23, 
              nbchicks = nbchicks, 
              nbpairs = nbpairs, 
              cov = (temp - mean(temp)) / sd(temp))
```
]

---

## Load R2jags to run Jags through R

.tiny-font[
```{r}
storks_temp <- jags(data  = datax,
               inits = inits,
               parameters.to.save = parameters,
               model.file = here::here("slides", "code", "logtemp.txt"),
               n.chains = 2,
               n.iter = nb.iterations,
               n.burnin = nb.burnin)
```
]

---

## Compute wAIC

```{r}
samples <- jags.samples(storks_temp$model,c("WAIC","deviance"), 
                        type = "mean", 
						n.iter = 2000,
						n.burnin = 1000,
						n.thin = 1)
samples$p_waic <- samples$WAIC
samples$waic <- samples$deviance + samples$p_waic
tmp <- sapply(samples, sum)
waic_temp <- round(c(waic = tmp[["waic"]], 
                     p_waic = tmp[["p_waic"]]),1)
```

---

## Model with rainfall only

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# read in data
datax <- list(N = 23, 
              nbchicks = nbchicks, 
              nbpairs = nbpairs, 
              cov = (rain - mean(rain)) / sd(rain))
```

---
## Load R2jags to run Jags through R

.tiny-font[
```{r}
storks_temp <- jags(data  = datax,
               inits = inits,
               parameters.to.save = parameters,
               model.file = here::here("slides", "code", "logtemp.txt"),
               n.chains = 2,
               n.iter = nb.iterations,
               n.burnin = nb.burnin)
```
]

---

## Compute wAIC

```{r}
samples <- jags.samples(storks_temp$model,c("WAIC","deviance"), 
                        type = "mean", 
						n.iter = 2000,
						n.burnin = 1000,
						n.thin = 1)
samples$p_waic <- samples$WAIC
samples$waic <- samples$deviance + samples$p_waic
tmp <- sapply(samples, sum)
waic_rain <- round(c(waic = tmp[["waic"]], 
                     p_waic = tmp[["p_waic"]]),1)
```

---
## Model with no effect of covariates

.tiny-font[
```{r}
# model specification
model <- 
paste("
model
{
	for( i in 1 : N) 
		{
		nbchicks[i] ~ dbin(p[i],nbpairs[i])
		logit(p[i]) <- a
		}
			
# priors for regression parameters
a ~ dnorm(0,0.001)
	}
")
writeLines(model,here::here("slides", "code", "lognull.txt"))
```
]

---
## MCMC details and data

.small-font[
```{r}
# list of lists of initial values (one for each MCMC chain)
init1 <- list(a = -0.5)
init2 <- list(a = 0.5)
inits <- list(init1,init2)
# specify parameters that need to be estimated
parameters <- c("a")
# specify nb iterations for burn-in and final inference 
nb.burnin <- 1000
nb.iterations <- 2000

# read in data
datax <- list(N = 23, 
              nbchicks = nbchicks, 
              nbpairs = nbpairs)
```
]

---
## Load R2jags to run Jags through R

.tiny-font[
```{r}
storks_temp <- jags(data  = datax,
               inits = inits,
               parameters.to.save = parameters,
               model.file = here::here("slides", "code", "lognull.txt"),
               n.chains = 2,
               n.iter = nb.iterations,
               n.burnin = nb.burnin)
```
]

---
## Compute wAIC

```{r}
samples <- jags.samples(storks_temp$model,c("WAIC","deviance"), 
                        type = "mean", 
						n.iter = 2000,
						n.burnin = 1000,
						n.thin = 1)
samples$p_waic <- samples$WAIC
samples$waic <- samples$deviance + samples$p_waic
tmp <- sapply(samples, sum)
waic_null <- round(c(waic = tmp[["waic"]], 
                     p_waic = tmp[["p_waic"]]),1)
```

---
## Compare WAIC

.tiny-font[
```{r}
data.frame(model = c('both_covariates', 'temp', 'rain', 'none'),
           waic = c(waic[1], waic_temp[1], waic_rain[1], waic_null[1]),
           p_waic = c(waic[2], waic_temp[2], waic_rain[2], waic_null[2])) %>%
  arrange(waic)
```
]

* Model with rainfall only seems to be better supported by the data. 

* In case models have similar wAIC values, model-averaging might be useful. 
