---
title: "Practical 3"
author: "Olivier Gimenez"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE}
#options(htmltools.dir.version = FALSE)
#knitr::opts_chunk$set(comment = "", warning = FALSE, message = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(palmerpenguins)
library(tidyverse)
```

## Moment matching

* In the capture-recapture example, we know a priori that the mean of the probability we're interested in is $\mu = 0.57$ and its variance is $\sigma^2 = 0.073^2$.

* Parameters $\mu$ and $\sigma^2$ are seen as the moments of a $Beta(\alpha,\beta)$ distribution.

* Now we look for values of $\alpha$ and $\beta$ that match the observed moments of the $Beta$ distribution.

* We need another set of equations:

$$\alpha = \bigg(\frac{1-\mu}{\sigma^2}- \frac{1}{\mu} \bigg)\mu^2$$

$$\beta = \alpha \bigg(\frac{1}{\mu}-1\bigg)$$

---
## Moment matching

* For our model, that means:

.small-font[
```{r echo = TRUE}
(alpha <- ( (1 - 0.57)/(0.073*0.073) - (1/0.57) )*0.57^2)
(beta <- alpha * ( (1/0.57) - 1))
```
]


---
## Question

* Use simulations to check that our estimates are correct. 

* Hint: The `R` function `rbeta()` might be of use.


---
# Solution

```{r eval = TRUE, echo = TRUE}
alpha <- ( (1 - 0.57)/(0.073*0.073) - (1/0.57) )*0.57^2
beta <- alpha * ( (1/0.57) - 1)
n <- 10000
samp <- rbeta(n, alpha, beta)
```

```{r eval = TRUE, echo = TRUE}
(mu <- mean(samp))
(sigma <- sqrt(var(samp)))
```

